---
title: "COVID-19 Metabolic Modeling Flux Analysis"
author: "Lillian Dillard"
date: "8/02/2021"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("~/")
library(vegan)
library(randomForest)
library(randomForestExplainer)
library(dplyr)
library(ggplot2)
library(readr)
library(readxl)
library(tidyverse)
library(xlsx)
library(UpSetR)
library(reactable)

# colorblind friendly palette with grey
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2",
               "#D55E00", "#CC79A7")
```

# COVID-19 Metabolomics Analysis

NMDS Analysis of ALL metabolites
```{r Metabolomics_NMDS, echo=TRUE}
#load data and set input and responses
model <- read_csv("endogenous_metabolomics.csv")

model$group <- as.factor(model$group)

#add value to each cell so no negatives
response <- model$group
input <- as.data.frame(data.matrix(model[-1])) + 1000

#run ordination with bray curtis
ord <- metaMDS(input)
data.scores <- as.data.frame(scores(ord))
data.scores$model <- response 

# Measure of centroid
model_centroids <- aggregate(cbind(NMDS1,NMDS2) ~ model, data = data.scores, mean)

#PERMANOVA
adon.results<-adonis(input ~ response, method="bray",perm=999)
print(adon.results)

#plot NMDS with measure of centroid for each model

p <- ggplot(data=data.scores) + 
  stat_ellipse(aes(x=NMDS1,y=NMDS2,colour=model),level = 0.50) +
  geom_point(aes(x=NMDS1,y=NMDS2,colour=model),size=1) +
  geom_point(data = model_centroids, aes(x=NMDS1, y=NMDS2, colour=model,
                                          shape = model), size = 4) +
  theme_classic()+
  xlab(NULL)+ ylab(NULL) +
  scale_color_manual(values = cbPalette,name = "Model Type",
                    labels = c("Non-Acute COVID-19",
                               "Severe COVID-19")) +
  scale_shape(guide = 'none')

p <- p + ggtitle("NMDS: Endogenous Differential Metabolomics")
p <- p + theme(text = element_text(size = 15))
p <- p + annotate(geom = 'text', label = 'R²= 0.11\np-value= 0.001', x = Inf, y = -.00025, hjust = 1, vjust = 0)

p
#right click graph and save as image
```

Random Forest to Select for MOST predictive metabolites
```{r Metabolomics_RF, echo=TRUE}
#convert all columns to names for random forest
names(model) <- make.names(names(model))

#conduct random forest with 500 trees
#https://cran.rstudio.com/web/packages/randomForestExplainer/vignettes/randomForestExplainer.html
set.seed(2017)
forest <- randomForest(group ~ ., data = model, localImp = TRUE)
importance_frame <- measure_importance(forest)
```

NMDS with only SIGNFICANTLY predictive Metabolites as identified by RF

```{r Metabolomics_nmds_filtered, echo=TRUE}

#run ordination with bray curtis
top_importance_frame <- importance_frame %>% 
  filter(p_value<=0.05) %>%
  filter(accuracy_decrease >= 0.0001) %>%
  select(c(variable))

#add value to each cell so no negatives
input <- as.data.frame(data.matrix(model[top_importance_frame$variable]))
response <- model$group

ord <- metaMDS(input)
data.scores <- as.data.frame(scores(ord))
data.scores$model <- response 

# Measure of centroid
model_centroids <- aggregate(cbind(NMDS1,NMDS2) ~ model, data = data.scores, mean)

#PERMANOVA for separation
adon.results<-adonis(input ~ response, method="bray",perm=999)
print(adon.results)

#plot NMDS with measure of centroid for each model

p <- ggplot(data=data.scores) + 
  stat_ellipse(aes(x=NMDS1,y=NMDS2,colour=model),level = 0.50) +
  geom_point(aes(x=NMDS1,y=NMDS2,colour=model),size=1) +
  geom_point(data = model_centroids, aes(x=NMDS1, y=NMDS2, colour=model,
                                          shape = model), size = 4) +
  theme_classic()+
  xlab(NULL)+ ylab(NULL) +
  scale_color_manual(values = cbPalette,name = "Model Type",
                    labels = c("Non-Acute COVID-19",
                               "Severe COVID-19")) +
  scale_shape(guide = 'none')

p <- p + ggtitle("Endogenous Differential Metabolomics")
p <- p + theme(text = element_text(size = 15))
p <- p + annotate(geom = 'text', label = 'R²= 0.18\np-value= 0.001', x = Inf, y = -.0035, hjust = 1, vjust = 0)

p
```
Upset Plot Comparing Models

```{r}
#Create UpsetR plot from nonacute, severe, open exchange lists from python. These reactions are present in each model but don't belong to consensus set across 4 models
listInput <- list(NonAcute = c('ALDD20xm', 'ALDD2xm', 'ALDD2y', 'AMPDA', 'EX_ps_hs_e', 'LCADi_D', 'NABTNO', 'PGLYCt', 'PVD3', 'PYNP2r', 'TMABADH', 'EX_ahdt_e', 'EX_HC00250_e', 'EX_HC01361_e', 'SPTc', 'r0242', 'r0249', 'GMPR', 'r0408', 'r0409', 'BETALDHx', 'AASAD3', 'r0610', 'r0642', 'r0688', 'r0892', 'r0940', 'RE3346M', 'INSTt4_2', 'PIt9', 'FAOXC15NADx', 'GLCt4', 'CYSTGL', 'DURIPP', 'GK1', 'SK_vitd3_c', 'PIt2r', 'RPE', 'TKT1', 'GCALDD', 'ALDD2x', 'SK_dchac_c', 'PFK', 'GLYALDDr', 'EX_pyr_e', 'IMACTD', 'LCADi', 'HMR_4079', 'URIK2', 'AGTix', 'ALDD20x', 'ALDD21', 'BAMPPALDOX', 'CYTDK1', 'DHCR71r', 'DHCR72r', 'EX_atp_e', 'EX_inost_e', 'EX_pglyc_hs_e', 'EX_utp_e', 'RE2112C', 'RE3346C', 'EX_crm_hs_e', 'PIt8', 'RE3012C', 'RE3012M', 'RE3012R', 'RPEc', 'DGK1', 'RPI', 'TKT2', 'DCMPDA', 'SK_cholate_c', 'ABUTD', 'CRMte', 'HMR_1565', 'LALDO2', 'HMR_4782', 'SK_ile__L_c', 'SK_leu__L_c', 'SK_trp__L_c', 'SK_glu__L_c', 'GCALDDm', 'GLACO', 'GLYCLTDy', 'HPYRRy', 'LCADm', 'PSt3', 'SPTix', 'TRIOK', 'URIK1', 'VD3', 'r0191', 'r0193', 'r0393', 'DUTCP', 'BETALDHy', 'r0587', 'r0611', 'r0643', 'r0707', '5HOXINDACTOX', 'r1116', 'DUTUP', 'FAOXC15NADPx', 'HMR_4816'),
                  Severe= c('ALCD21_D', 'ALCD22_D', 'ALDD20xm', 'ALDD2xm', 'ALDD2y', 'AMPDA', 'FALDH', 'HPYRR2x_1', 'LALDO', 'LCADi_D', 'NABTNO', 'PVD3', 'PYNP2r', 'TMABADH', 'SPTc', 'r0249', 'GMPR', 'r0408', 'r0409', 'BETALDHx', 'AASAD3', 'r0610', 'r0642', 'r0688', 'r1378', 'RE3346M', 'FAOXC15NADx', 'DURIPP', 'GK1', 'SK_vitd3_c', 'RPE', 'TKT1', 'GCALDD', 'ALDD2x', 'SK_dchac_c', 'PFK', 'GLYALDDr', 'EX_pyr_e', 'IMACTD', 'LCADi', 'HMR_4079', 'HMR_6633', 'HMR_9804', '34DHOXPEGOX', 'AGTix', 'LCARS', 'ALCD22_L', 'ALCD2x', 'ALDD20x', 'ALDD21', 'BAMPPALDOX', 'DHCR71r', 'DHCR72r', 'RE2112C', 'RE3346C', 'RE3012C', 'RE3012M', 'RE3012R', 'RPEc', 'SK_akg_c', 'DGK1', 'RPI', 'TKT2', 'DCMPDA', 'SK_cholate_c', 'ALCD1', 'ABUTD', 'GLXO1', 'LDH_D', 'HMR_1129', 'HMR_1565', 'LALDO2', 'HMR_3859', 'SK_glu__L_c', 'GCALDDm', 'GLACO', 'GLYCLTDy', 'HPYRRy', 'LCADm', 'LDH_Lm', 'SPTix', 'TRIOK', 'VD3', 'r0191', 'r0393', 'BETALDHy', 'r0611', 'r0643', 'r0739', '5HOXINDACTOX', 'FAOXC15NADPx'),
                  Open = c('ALCD21_D', 'ALCD22_D', 'EX_ps_hs_e', 'FALDH', 'HPYRR2x_1', 'LALDO', 'PGLYCt', 'EX_ahdt_e', 'EX_HC00250_e', 'EX_HC01361_e', 'r0242', 'r0892', 'r0940', 'r1378', 'INSTt4_2', 'PIt9', 'GLCt4', 'CYSTGL', 'PIt2r', 'HMR_6633', 'URIK2', 'HMR_9804', '34DHOXPEGOX', 'LCARS', 'ALCD22_L', 'ALCD2x', 'CYTDK1', 'EX_atp_e', 'EX_inost_e', 'EX_pglyc_hs_e', 'EX_utp_e', 'EX_crm_hs_e', 'PIt8', 'SK_akg_c', 'ALCD1', 'GLXO1', 'LDH_D', 'CRMte', 'HMR_1129', 'HMR_3859', 'HMR_4782', 'SK_ile__L_c', 'SK_leu__L_c', 'SK_trp__L_c', 'LDH_Lm', 'PSt3', 'URIK1', 'r0193', 'DUTCP', 'r0587', 'r0707', 'r0739', 'r1116', 'DUTUP', 'HMR_4816'))
names(listInput) = c("Non-Acute COVID-19", "Severe COVID-19", "Open Exchanges")
upsetplot <-upset(fromList(listInput), 
                  mainbar.y.label='Number of Shared Reactions',
                  sets.x.label = 'Nonconsensus Rxns',
                  text.scale = c(1.8,1,1.5,.8),
                  point.size = 5, 
                  line.size = 1.2,
                  order.by = "freq",
                  keep.order = TRUE)
upsetplot #manually save from viewer so that text fits
```

# Model 1: open model vs min media model(nonacute) vs severe model

Conduct Random Forest: identify most predictive reactions for each model

Definitions

Mean Minimal Depth: Minimal depth for a variable in a tree equals to the depth 
                    of the node which splits on that variable and is the closest 
                    to the root of the tree
Times a Root: total number of trees in which Xj is used for splitting the root 
              node (i.e., the whole sample is divided into two based on the 
              value of Xj)

Number of Nodes: total number of nodes that use Xj for splitting (it is usually 
                equal to no_of_trees if trees are shallow)

```{r Model1_RF, echo=TRUE}
#load data and convert rxn ids to rxn names 
model1 <- read_excel("commonrxn_fluxdata.xlsx", sheet = 1)
model1$Model <- as.factor(model1$Model)

#add pre-fix to column names so that none start with number --> will cause issue with random forest
colnames(model1) <- paste("pre", colnames(model1), sep = "_")

#conduct random forest with 500 trees
#https://cran.rstudio.com/web/packages/randomForestExplainer/vignettes/randomForestExplainer.html
set.seed(2017)
forest <- randomForest(pre_Model ~ ., data = model1, localImp = TRUE)
varImpPlot(forest, n.var = 10, main = "Top 10 Predictive Rxns")

#https://cran.rstudio.com/web/packages/randomForestExplainer/vignettes/randomForestExplainer.html
importance_frame <- measure_importance(forest)
plot_multi_way_importance(importance_frame, size_measure = "no_of_nodes") + 
  theme_classic()
plot_importance_rankings(importance_frame)
```

Create NMDS Plot & run statistics to assess model separation

```{r Model1_NMDS, echo=TRUE}
#load data and set input and responses
model1 <- read_excel("commonrxn_fluxdata.xlsx", sheet = 1)
model1$Model <- as.factor(model1$Model)

#add value to each cell so no negatives
response <- model1$Model
input <- model1[-1] + 1000

#run ordination with bray curtis
ord <- metaMDS(input)
data.scores <- as.data.frame(scores(ord))
data.scores$model <- response 

# Measure of centroid
model1_centroids <- aggregate(cbind(NMDS1,NMDS2) ~ response, data = data.scores, mean)

#PERMANOVA
adon.results<-adonis(input ~ response, method="bray",perm=999)
print(adon.results)

#plot NMDS with measure of centroid for each model

p <- ggplot(data=data.scores) + 
  stat_ellipse(aes(x=NMDS1,y=NMDS2,colour=response),level = 0.50) +
  geom_point(aes(x=NMDS1,y=NMDS2,shape=response,colour=response),size=1) +
  geom_point(data = model1_centroids, aes(x=NMDS1, y=NMDS2, colour=response,
                                          shape = response), size = 4) +
  theme_classic()+
  xlab(NULL)+ ylab(NULL) +
  scale_color_manual(values = cbPalette,name = "Model Type",
                     labels = c("Non-Acute COVID-19 Model",
                                "Open Exchange Model",
                                "Severe COVID-19 Model")) +
  scale_shape(guide = 'none')
p <- p + ggtitle("Differential Core Metabolism")
p <- p + theme(text = element_text(size = 15))
p <- p + annotate(geom = 'text', label = 'R²= 0.42\np-value= 0.001', x = Inf, y = -.075, hjust = 1, vjust = 0)

p
```

# Model 2: severe vs non-acute model 

```{r Model2_RF, echo=TRUE}
#load data and set input and responses
model2 <- read_excel("commonrxn_fluxdata.xlsx", sheet = 2)
model2$Model <- as.factor(model2$Model)

#add pre-fix to column names so that none start with number --> will cause issue with random forest
colnames(model2) <- paste("pre", colnames(model2), sep = "_")

#conduct random forest with 500 trees
#https://cran.rstudio.com/web/packages/randomForestExplainer/vignettes/randomForestExplainer.html
set.seed(2017)
forest2 <- randomForest(pre_Model ~ ., data = model2, localImp = TRUE)
importance_frame2 <- measure_importance(forest2)
varImpPlot(forest2, n.var = 10, main = "Top 10 Predictive Rxns")

plot_multi_way_importance(importance_frame2, size_measure = "no_of_nodes")
plot_importance_rankings(importance_frame2)
```


```{r Model2_NMDS, echo=TRUE}
#load data and set input and responses
model2 <- read_excel("commonrxn_fluxdata.xlsx", sheet = 2)
model2$Model <- as.factor(model2$Model)

#add value to each cell so no negatives
response2 <- model2$Model
input2 <- model2[-1] + 1000

#run ordination with bray curtis
ord2 <- metaMDS(input2)
data.scores2 <- as.data.frame(scores(ord2))
data.scores2$model <- response2 

# Measure of centroid
model2_centroids <- aggregate(cbind(NMDS1,NMDS2) ~ model, data = data.scores2, mean)

#PERMANOVA
adon.results2<-adonis(input2 ~ response2, method="bray",perm=999)
print(adon.results2)

#plot NMDS with measure of centroid for each model
p <- ggplot(data=data.scores2) + 
  stat_ellipse(aes(x=NMDS1,y=NMDS2,colour=model),level = 0.50) +
  geom_point(aes(x=NMDS1,y=NMDS2,shape=model,colour=model),size=1) +
  geom_point(data = model2_centroids, aes(x=NMDS1, y=NMDS2, colour=model,
                                          shape = model), size = 4) +
  theme_classic()+
  xlab(NULL)+ ylab(NULL) +
  scale_color_manual(values = cbPalette,name = "Model Type",
                     labels = c("Non-Acute COVID-19 Model",
                                "Severe COVID-19 Model")) +
  scale_shape(guide = 'none')
p <- p + ggtitle("Flux Shifts between Severe vs. Non-Acute")
p <- p + theme(text = element_text(size = 15))
p <- p + annotate(geom = 'text', label = 'R²= 0.42\np-value= 0.001', x = Inf, y = -.075, hjust = 1, vjust = 0)
p
```